---
title: "Yates' Continuity Correction"
subtitle: "Assessment of Utility"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10)
```

## Introduction

The `tbl_summary() %>% add_p()` function from the gtsummary package is updating the default test for most continuous variables.
Previously, continuous variables being compared between two groups would be compared with `chisq.test(correct = TRUE)`, which would use a continuity correction for 2 by 2 table comparisons.
However, the simulation study below demonstrates the superior properties of `chisq.test(correct = FALSE)`, and we have updated the default to use this test instead.

## Simulating results

We will simulate many studies assuming various prevalence rates of two binary variables and at various sample sizes.
The power and size of the chi-squared tests will be reported.

Attach needed libraries.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(gtsummary)
```

Set the number of simulated studies that will be run for each combination of rates and sample size.

```{r, warning=FALSE}
simn <- 100
```

Simulate the trial results.

```{r, warning=FALSE}
sim_results <-
  list(
    simn = seq_len(simn),                              # number of simulations per scenario
    n = c(20, 50, 100, 500),                           # sample sizes
    p = c(0.05, 0.10, 0.20, 0.50),                     # rate of success in group 1
    delta = c(0.0, 0.05, 0.10, 0.20, 0.30, 0.40, 0.45) # p + delta = rate of success in group 2
  ) %>%
  cross_df() %>%
  # deleting duplicated scenarios (due to symmetry of test)
  filter(p + delta <= 0.5) %>%
  mutate(
    # simulating  data set with specified success rates, and calculating p-values
    df_result = pmap(
      list(n, p, delta),
      function(n, p, delta) {
        data = 
          bind_rows(
            tibble(value = runif(n/2) < p, group = 1),
            tibble(value = runif(n/2) < p + delta, group = 2),
          )
        
        tribble(
          ~test, ~reject, 
          "With Correction", with(data, table(value, group)) %>% 
            {tryCatch(chisq.test(., correct = TRUE), error = function(e) NA)$p.value} %>% 
            {. < 0.05},
          "Without Correction", with(data, table(value, group)) %>% 
            {tryCatch(chisq.test(., correct = FALSE), error = function(e) NA)$p.value} %>% 
            {. < 0.05}
        )
      }
    )
  ) %>%
  unnest(df_result) %>% 
  # deleting sets of results where at least one of the p-values are missing
  group_by(simn, n, p, delta) %>% 
  filter(sum(is.na(reject)) == 0) %>%
  ungroup()
```

Summarize simulation results.

```{r, warning=FALSE}
# calculating power (or level) of test across simulations
results <- 
  sim_results %>%
  group_by(n, p, delta, test) %>%
  summarise(power = mean(reject),
            simn = n(),
            .groups = "drop") 
```

## Power of test

Across sample sizes and various prevalence of the binary variables, the test without continuity correction exhibits higher power.

```{r, fig.height = 7.5}
results %>%
  filter(p != 0.5) %>%
  mutate(n = str_glue("N = {n}") %>% factor(levels = unique(.)), 
         p = str_glue("{p * 100}%") %>% factor(levels = unique(.))) %>%
  ggplot(aes(x = delta, y = power, color = test)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.05) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_grid(n ~ p, scale = "free") +
  labs(color = "",
       y = "Power",
       x = "Delta") +
  theme(legend.position="bottom")
```

## Level of test

In the simulation study, we rejected the null hypothesis when the p-value was less than 0.05.
The true rate of rejection should be 5%, and the test without continuity correction is closer to this threshold.

```{r, fig.height = 5}
results %>%
  filter(delta == 0) %>%
  mutate(n = str_glue("N = {n}") %>% factor(levels = unique(.))) %>%
  ggplot(aes(x = p, y = power, color = test)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~n) +
  geom_hline(yintercept = 0.05) +
  labs(color = "",
       y = "Prob. of Rejecting H0",
       x = "Event Rate") +
  theme(legend.position="bottom")
```

## Restore Previous Behavior

If a user wishes to continue using the continuity correction, there are two options.

1. Specify the variables you'd like to compare with the chi-squared test without continuity correction. For example, this line of code specifies all categorical variables are compared with `"chisq.test"` that uses the continuity correction: `trial %>% tbl_summary(by = trt) %>% add_p(all_categorical() ~ "chisq.test")`

1. If you want to make the global change for the entire script or R session, use the  [gtsummary themes](http://www.danieldsjoberg.com/gtsummary/articles/themes.html). Run the following line anywhere in your R script, which will make the chi-sequared test with continuity correction the default: `set_gtsummary_theme(list("add_p.tbl_summary-attr:test.continuous_by2" = "chisq.test"))`